Pattern matching for\\ $(213,231)$-avoiding permutations

### 2015-04-19 article.tex

\begin{corollary}
\label{corollary:max_is_ascent}
For $\pi \in \AV_n(213,231)$ if $\pi[i]$ is an ascent element then $\pi[i]$ is the minimal of $\pi[i:]$
and if if $\pi[i]$ is a descent element then $\pi[i]$ is the maximal of $\pi[i:]$
\end{corollary}

@@

This section focuses on the pattern containment problem
in case only the pattern $\sigma$ avoids $231$ and $213$.
We need to consider a specific decomposition $\factora$ of $\sigma$ into factors.
For $\sigma \in S_k$, the factorization
$\sigma = \factor{\pmotif}{p} \factor{\pmotif}{p-1} \ldots \factor{\pmotif}{1}$ is defined as follows:
(i) $\sigma[k]$ is the last letter of $\factor{\pmotif}{1}$,
(ii)for every $1 \leq i < k$,
if $s_\sigma[i] = s_\sigma[i+1]$ then
$\sigma[i]\sigma[i+1]$ is part of the same factor.
For every factor $\factor{\pmotif}{j}$ of the decomposition, we let
$\first{\pmotif}{j}$ stand for the leftmost letter of $\factor{\pmotif}{j}$,
and $\firsti{\pmotif}{j}$ stand for the index of the leftmost letter of $\factor{\pmotif}{j}$.
For example,
for $\sigma = 981237654$,
we have
$s_\sigma = ddaaaddd$ and hence
$\sigma = \factor{\pmotif}{3} \factor{\pmotif}{2} \factor{\pmotif}{1}$ with
$\factor{\pmotif}{1} =7654$, $\factor{\pmotif}{2} = 123$ and $\factor{\pmotif}{3} = 98$.
Furthermore,
$\first{\pmotif}{1} = 7$, $\first{\pmotif}{2} = 1$, $\first{\pmotif}{3} = 9$,
$\firsti{\pmotif}{1} = 6$, $\firsti{\pmotif}{2} = 3$ and $\firsti{\pmotif}{3} = 1$.

@@

\begin{proof}
Take the right most element of $\factor{\sigma}{i}$, if this element is an ascent element,
then it is smaller than the next element which is $\first{\sigma}{i-1}$. Because $\factor{\sigma}{i}$ is an increasing subsequence, every element of $\factor{\sigma}{i}$ is inferior to $\first{\sigma}{i-1}$.
By lemma \ref{lemma:first element is 1 or n}, $\first{\sigma}{i-1}$ is the maximal element of $\factor{\sigma}{i-1}\ldots\factor{\sigma}{1}$ thus $\first{\sigma}{i-1}$ is the maximal element of $\factor{\sigma}{i}\factor{\sigma}{i-1}\ldots\factor{\sigma}{1}$.
We use a similar argument if the rightmost element is a decent element.
$\qed$
\end{proof}

To ease the next lemma and its proof we define the set $\SET{\pmotif}{\ptext}{i}{j}$ such that if $s \in S$, then $s$ is an occurrence of $\pmotif[\firsti{\pmotif}{i}:]$ in $\ptext[j:]$ and the element matching $\first{\pmotif}{i}$ is $\ptext[j]$.

@@

Consider the following problem :\\
$
\LM{\pmotif}{\ptext}{i}{j} = 
\begin{cases}
	\text{The maximal value of a subsequence} 
		& \text{If $\first{\sigma}{i}$ is  } \\
	\text{ which is an occurence of $\sigma[\firsti{\sigma}{i}:]$ in $\pi[j:]$} & 
		\text{ an ascent element } \\	
	\text{ and that minimize the maximal value} & \\
	\text{ and $\first{\sigma}{i}$ is matched with $\ptext[j]$} & \\

@@

\textbf{STEP :} \\
For clarity of reading we consider 3 cases for the problem $\PM{\pbmotif}{\ptext}{\lb}{\ub}{i}{j}$ :
\begin{itemize}
	\item If $\ptext[j] \notin [\lb,\ub]$ then :
	$$
	\PM{\pbmotif}{\ptext}{\lb}{\ub}{i}{j} = false
	$$
	this is straightforward with the definition. If $\ptext[j] \notin [\lb,\ub]$ then it can not be part of an occurrence of  $\pmotif[i:]$ in $\ptext[i:]$ with every element is in $[\lb,\ub]$.

@@

	\end{cases}
	$$	
	$\pmotif[i]$ can occur in $\ptext[j]$ because $\ptext[j] \in [\lb,\ub]$. $\pmotif[i:]$ occurs in $\ptext[j:]$, if and only if $\pmotif[i+1:]$ occurs in $\ptext[j+1:]$. $\pmotif[i+1:]$ occurs in $\ptext[j+1:]$ if and only if there is an occurrence of $\pmotif[i+1:]$ in $\ptext[j+1:]$ starting at element $k$, $j<k$, which correspond to know if $\bigcup_{j<k} \PM{\pbmotif}{\ptext}{*}{*}{i+1}{k}$ is true. Because $\pmotif[i]$ is an ascent element,  $\pmotif[i]$ is the minimal element of $\pmotif[i:]$, thus a match of $\pmotif[i+1:]$ can not use value smaller than $\ptext[j]$, so $\ptext[j]+1$ is a (new) lower bound. The upper bound is unchanged. So there is a occurrence of  $\pmotif[i:]$ starting at element $j$ in $\ptext[j:]$ if and only if $\bigcup_{j<k} \PM{\pbmotif}{\ptext}{\ptext[j]+1}{\ub}{i+1}{k}$.

	\item If $\ptext[j] \in [\lb,\ub]$ and $\pmotif[i]$ is a descent element then :
	$$
	\PM{\pbmotif}{\ptext}{\lb}{\ub}{i}{j}=
	\begin{cases}
			\bigcup_{j<k} \PM{\pbmotif}{\ptext}{\lb}{\ptext[j]-1}{i+1}{k}
				& \text{if $\pmotif[i]$ is not underlined } \\
				& \text{ and $\pmotif[i]$ is not overlined} \\ 
			\bigcup_{j<k} \PM{\pbmotif}{\ptext}{\lb}{\ptext[j]-1}{i+1}{k}
				& \text{if $\pmotif[i]$ is not underlined } \\
				& \text{ and $\overline{\pmotif[i](\pmotif[i]+1)}$ or ${\pmotif[i]}^\urcorner$}\\ 	
				& \text{ and $\ptext[j]=\ub$} \\
			\PM{\pbmotif}{\ptext}{\lb}{\ptext[j]-1}{i+1}{j+1}	
				& \text{if $\underline{\pmotif[i]\pmotif[i+1]}$ } \\
				& \text{ and $\overline{\pmotif[i](\pmotif[i]+1)}$ or ${\pmotif[i]}^\urcorner$}\\ 	
				& \text{ and $\ptext[j]=\ub$} \\					
			\PM{\pbmotif}{\ptext}{\lb}{\ptext[j]-1}{i+1}{j+1}	
				& \text{if $\underline{\pmotif[i]\pmotif[i+1]}$ } \\
				& \text{ and $\pmotif[i]$ is not overlined} \\	
			false & \text{otherwise}		
	\end{cases}
	$$	
	We prove symmetrically as the last case that 
	$\pmotif[i:]$ occurs in $\ptext[j:]$
	with $\sigma[i]$ occurs in $\pi[j]$.
	 starting at element $j$ if and only if $\bigcup_{j<k} \PM{\pbmotif}{\ptext}{\lb}{\ptext[j]-1}{i+1}{k}$.

@@

\paragraph{Value Constraint.} There are 3 types of position constraints 
\begin{itemize}
	\item If $^\ulcorner{\sigma[i]}$ (and thus $\sigma[i]=1$) then the minimal value of $\pmotif$ must occur in the minimal value of $\ptext$. 
	\begin{itemize}
		\item If $\sigma[i]$ is an ascent element then $\sigma[i]$ is the left most ascent element (if not $\sigma[i]$ would not be the minimal element, because ascent elements are increasing). Remark that the problem starts with the lower bound as $1$, and inductive problems before the index $i$ are called with descent elements which do not modified the lower bound, so for every problem with index $i$ has for lower bound $1$. 
		$\PM{\pbmotif}{\ptext}{1}{*}{i}{*}$ is either an inductive case or a base case, in either way  $\PM{\pbmotif}{\ptext}{1}{*}{i}{*}$ is true if and only if the occurrence of $\sigma[i]$ is equal to lower bound which is  $1$.

		\item If $\sigma[i]$ is a descent element then $i=n_\pmotif$ ($\sigma[i]$ is the right most element). Which is solved in the case case.
	\end{itemize}

	\item If ${\pmotif[i]}^\urcorner$ (and thus $\sigma[i]=n_\pmotif$) then the maximal value of $\pmotif$ must occur in the maximal value of $\ptext$.
	\begin{itemize}
		\item If $\sigma[i]$ is a descent element then $\sigma[i]$ is the left most ascent element (if not $\sigma[i]$ would not be the maximal element, because descent elements are decreasing). Remark that the problem starts with the upper bound as $n_\ptext$, and any inductive problems before the index $i$ are called with ascent elements which do not modified the upper bound, every problem with index $i$ has for upper bound $n_\ptext$. 		$\PM{\pbmotif}{\ptext}{*}{n_\pmotif}{i}{*}$ is either an inductive case or a base case, in either way  $\PM{\pbmotif}{\ptext}{*}{n_\pmotif}{i}{*}$ is true if and only if the occurrence of $\sigma[i]$ is equal to upper bound which is  $n_\pmotif$.

		\item If $\sigma[i]$ is an ascent element then $\sigma[i]$ then $i=n_\pmotif$ ($\sigma[i]$ is the right most element) which is solved in the base case.
	\end{itemize}

	\item  If $\overline{\pmotif[i]\pmotif[i']}$, with $\pmotif[i']=\pmotif[i]+1$ then if $\pmotif[i]$ occurs in $\ptext[j]$ then $\pmotif[i']$ must occur in $\ptext[j]+1$. 
		\begin{itemize}
			\item The case $\pmotif[i]$ is a descent element, $\pmotif[i']$ is an ascent element and $i<i'$ (remark that this case is equivalent to the case $\pmotif[i]$ is an ascent element, $\pmotif[i']$ is a descent element and $i'<i$) is not possible. 
			Indeed $\pmotif[i]$ is the maximal element of $\pmotif[i:]$ thus $\pmotif[i] > \pmotif[i']$ which is in contradiction with 
			$\pmotif[i']=\pmotif[i]+1$.

@@

			\item If $\pmotif[i]$ is an ascent element and $\pmotif[i']$ is an ascent element then $i<i'$ and there is no  ascent element between $\pmotif[i]$ and $\pmotif[i']$ (lemma \ref{lemma:ascentDescentAscent}), Thus inductive problems between $i+1$ and $i'$ do not modified the lower bound. Remark that if $\pmotif[i]$ occurs in $\ptext[j]$ then 
			the inductive problems after $\PM{\pbmotif}{\ptext}{*}{*}{i}{*}$ have for lower bound $\ptext[j]+1$. $\PM{\pbmotif}{\ptext}{*}{*}{i'}{*}$ is either an inductive case or a base case, in either way  $\PM{\pbmotif}{\ptext}{*}{*}{i'}{*}$ is true if and only if the occurrence of $\sigma[i']$ is equal to lower bound which is equal to $\sigma[i']+1$ so the occurrence of $\pmotif[i]$ and $\pmotif[i']$ are consecutive in value.
		\end{itemize}

>>>>

\begin{corollary}
\label{corollary:max is ascent}
For $\pi \in \AV_n(213,231)$ if $\pi[i]$ is an ascent element then $\pi[i]$ is the minimal of $\pi[i:]$
and if if $\pi[i]$ is a descent element then $\pi[i]$ is the maximal of $\pi[i:]$
\end{corollary}

@@

This section focuses on the pattern containment problem
in case only the pattern $\sigma$ avoids $231$ and $213$.
We need to consider a specific decomposition $\factora$ of $\sigma$ into factors.
For $\sigma \in S_k$, the \emph{factorization}
$\sigma = \factor{\pmotif}{p} \factor{\pmotif}{p-1} \ldots \factor{\pmotif}{1}$
is defined as follows:
(i) $\sigma[k]$ is the last letter of $\factor{\pmotif}{1}$,
(ii)for every $1 \leq i < k$,
if $s_\sigma[i] = s_\sigma[i+1]$ then
$\sigma[i]\sigma[i+1]$ is part of the same factor.
For every factor $\factor{\pmotif}{j}$ of the decomposition, we let
$\first{\pmotif}{j}$ stand for the leftmost letter of $\factor{\pmotif}{j}$,
and $\firsti{\pmotif}{j}$ stand for the index in $\sigma$
of the leftmost letter of $\factor{\pmotif}{j}$.
For example,
for $\sigma = 981237654$,
we have
$s_\sigma = ddaaaddd$ and hence
$\sigma = \factor{\pmotif}{3} \factor{\pmotif}{2} \factor{\pmotif}{1}$ with
$\factor{\pmotif}{1} =7654$, $\factor{\pmotif}{2} = 123$ and $\factor{\pmotif}{3} = 98$.
Furthermore,
$\first{\pmotif}{1} = 7$, $\first{\pmotif}{2} = 1$, $\first{\pmotif}{3} = 9$,
$\firsti{\pmotif}{1} = 6$, $\firsti{\pmotif}{2} = 3$ and $\firsti{\pmotif}{3} = 1$.

@@

\begin{proof}
Consider the rightmost element of $\factor{\sigma}{i}$ and assume this element is an ascent.
Then, it follows that it is smaller than the following element which is $\first{\sigma}{i-1}$.
Now, since $\factor{\sigma}{i}$ is an increasing subsequence, every element of
$\factor{\sigma}{i}$ is smaller than $\first{\sigma}{i-1}$.
According to Lemma\ref{lemma:first element is 1 or n}, $\first{\sigma}{i-1}$ is the maximal (minimal) element of $\factor{\sigma}{i-1}\ldots\factor{\sigma}{1}$, and hence
$\first{\sigma}{i-1}$ is the maximal  element of
$\factor{\sigma}{i}\factor{\sigma}{i-1}\ldots\factor{\sigma}{1}$.

A similar argument shows that the lemma holds true if the rightmost element
of $\factor{\sigma}{i}$ is a descent.
\qed
\end{proof}

For easy readability, we now define the set
$\SET{\pmotif}{\ptext}{i}{j}$ as follows:
if $s \in \SET{\pmotif}{\ptext}{i}{j}$ then
$s$ is an occurrence of $\pmotif[\firsti{\pmotif}{i}:]$ in $\ptext[j:]$
and the element matching $\first{\pmotif}{i}$ is $\ptext[j]$.

@@

Consider the following problem :\\
$
\LM{\pmotif}{\ptext}{i}{j} =
\begin{cases}

	\text{The maximal value of a subsequence} 
		& \text{If $\first{\sigma}{i}$ is  } \\
	\text{ which is an occurence of $\sigma[\firsti{\sigma}{i}:]$ in $\pi[j:]$} & 
		\text{ an ascent element } \\	
	\text{ and that minimize the maximal value} & \\
	\text{ and $\first{\sigma}{i}$ is matched with $\ptext[j]$} & \\

@@

\textbf{STEP :} \\
For clarity of reading we consider 3 cases for the problem $\PM{\pbmotif}{\ptext}{\lb}{\ub}{i}{j}$ :
\begin{itemize}
	\item If $\ptext[j] \notin [\lb,\ub]$ then :
	$$
	\PM{\pbmotif}{\ptext}{\lb}{\ub}{i}{j} = false
	$$

	this is straightforward with the definition. If $\ptext[j] \notin [\lb,\ub]$ then it can not be part of an occurrence of  $\pmotif[i:]$ in $\ptext[i:]$ with every element is in $[\lb,\ub]$.

@@

	\end{cases}
	$$

	$\pmotif[i]$ can occur in $\ptext[j]$ because $\ptext[j] \in [\lb,\ub]$. $\pmotif[i:]$ occurs in $\ptext[j:]$, if and only if $\pmotif[i+1:]$ occurs in $\ptext[j+1:]$. $\pmotif[i+1:]$ occurs in $\ptext[j+1:]$ if and only if there is an occurrence of $\pmotif[i+1:]$ in $\ptext[j+1:]$ starting at element $k$, $j<k$, which correspond to know if $\bigcup_{j<k} \PM{\pbmotif}{\ptext}{*}{*}{i+1}{k}$ is true. Because $\pmotif[i]$ is an ascent element,  $\pmotif[i]$ is the minimal element of $\pmotif[i:]$, thus a match of $\pmotif[i+1:]$ can not use value smaller than $\ptext[j]$, so $\ptext[j]+1$ is a (new) lower bound. The upper bound is unchanged. So there is a occurrence of  $\pmotif[i:]$ starting at element $j$ in $\ptext[j:]$ if and only if $\bigcup_{j<k} \PM{\pbmotif}{\ptext}{\ptext[j]+1}{\ub}{i+1}{k}$.

	\item If $\ptext[j] \in [\lb,\ub]$ and $\pmotif[i]$ is a descent element then :

	$$
	\PM{\pbmotif}{\ptext}{\lb}{\ub}{i}{j}=
	\begin{cases}
			\bigcup_{j<k} \PM{\pbmotif}{\ptext}{\lb}{\ptext[j]-1}{i+1}{k}
				& \text{if $\pmotif[i]$ is not underlined } \\
				& \text{ and $\pmotif[i]$ is not overlined} \\
			\bigcup_{j<k} \PM{\pbmotif}{\ptext}{\lb}{\ptext[j]-1}{i+1}{k}
				& \text{if $\pmotif[i]$ is not underlined } \\
				& \text{ and $\overline{\pmotif[i](\pmotif[i]+1)}$ or ${\pmotif[i]}^\urcorner$}\\
				& \text{ and $\ptext[j]=\ub$} \\
			\PM{\pbmotif}{\ptext}{\lb}{\ptext[j]-1}{i+1}{j+1}
				& \text{if $\underline{\pmotif[i]\pmotif[i+1]}$ } \\
				& \text{ and $\overline{\pmotif[i](\pmotif[i]+1)}$ or ${\pmotif[i]}^\urcorner$}\\
				& \text{ and $\ptext[j]=\ub$} \\
			\PM{\pbmotif}{\ptext}{\lb}{\ptext[j]-1}{i+1}{j+1}
				& \text{if $\underline{\pmotif[i]\pmotif[i+1]}$ } \\
				& \text{ and $\pmotif[i]$ is not overlined} \\
			false & \text{otherwise}
	\end{cases}
	$$	
	We prove symmetrically as the last case that 
	$\pmotif[i:]$ occurs in $\ptext[j:]$
	with $\sigma[i]$ occurs in $\pi[j]$.
	 starting at element $j$ if and only if $\bigcup_{j<k} \PM{\pbmotif}{\ptext}{\lb}{\ptext[j]-1}{i+1}{k}$.

@@

\paragraph{Value Constraint.} There are 3 types of position constraints
\begin{itemize}
	\item If $^\ulcorner{\sigma[i]}$ (and thus $\sigma[i]=1$) then the minimal value of $\pmotif$ must occur in the minimal value of $\ptext$.
	\begin{itemize}

		\item If $\sigma[i]$ is an ascent element then $\sigma[i]$ is the left most ascent element (if not $\sigma[i]$ would not be the minimal element, because ascent elements are increasing). Remark that the problem starts with the lower bound as $1$, and inductive problems before the index $i$ are called with descent elements which do not modified the lower bound, so for every problem with index $i$ has for lower bound $1$. 
		$\PM{\pbmotif}{\ptext}{1}{*}{i}{*}$ is either an inductive case or a base case, in either way  $\PM{\pbmotif}{\ptext}{1}{*}{i}{*}$ is true if and only if the occurrence of $\sigma[i]$ is equal to lower bound which is  $1$.

		\item If $\sigma[i]$ is a descent element then $i=n_\pmotif$ ($\sigma[i]$ is the right most element). Which is solved in the case case.

	\item If ${\pmotif[i]}^\urcorner$ (and thus $\sigma[i]=n_\pmotif$) then the maximal value of $\pmotif$ must occur in the maximal value of $\ptext$.
	\begin{itemize}

		\item If $\sigma[i]$ is a descent element then $\sigma[i]$ is the left most ascent element (if not $\sigma[i]$ would not be the maximal element, because descent elements are decreasing). Remark that the problem starts with the upper bound as $n_\ptext$, and any inductive problems before the index $i$ are called with ascent elements which do not modified the upper bound, every problem with index $i$ has for upper bound $n_\ptext$. 		$\PM{\pbmotif}{\ptext}{*}{n_\pmotif}{i}{*}$ is either an inductive case or a base case, in either way  $\PM{\pbmotif}{\ptext}{*}{n_\pmotif}{i}{*}$ is true if and only if the occurrence of $\sigma[i]$ is equal to upper bound which is  $n_\pmotif$.

		\item If $\sigma[i]$ is an ascent element then $\sigma[i]$ then $i=n_\pmotif$ ($\sigma[i]$ is the right most element) which is solved in the base case.

	\item  If $\overline{\pmotif[i]\pmotif[i']}$, with $\pmotif[i']=\pmotif[i]+1$ then if $\pmotif[i]$ occurs in $\ptext[j]$ then $\pmotif[i']$ must occur in $\ptext[j]+1$.
		\begin{itemize}

			\item The case $\pmotif[i]$ is a descent element, $\pmotif[i']$ is an ascent element and $i<i'$ (remark that this case is equivalent to the case $\pmotif[i]$ is an ascent element, $\pmotif[i']$ is a descent element and $i'<i$) is not possible. 
			Indeed $\pmotif[i]$ is the maximal element of $\pmotif[i:]$ thus $\pmotif[i] > \pmotif[i']$ which is in contradiction with 
			$\pmotif[i']=\pmotif[i]+1$.

@@

			\item If $\pmotif[i]$ is an ascent element and $\pmotif[i']$ is an ascent element then $i<i'$ and there is no  ascent element between $\pmotif[i]$ and $\pmotif[i']$ (lemma \ref{lemma:ascentDescentAscent}), Thus inductive problems between $i+1$ and $i'$ do not modified the lower bound. Remark that if $\pmotif[i]$ occurs in $\ptext[j]$ then

			the inductive problems after $\PM{\pbmotif}{\ptext}{*}{*}{i}{*}$ have for lower bound $\ptext[j]+1$. $\PM{\pbmotif}{\ptext}{*}{*}{i'}{*}$ is either an inductive case or a base case, in either way  $\PM{\pbmotif}{\ptext}{*}{*}{i'}{*}$ is true if and only if the occurrence of $\sigma[i']$ is equal to lower bound which is equal to $\sigma[i']+1$ so the occurrence of $\pmotif[i]$ and $\pmotif[i']$ are consecutive in value.
		\end{itemize}
